---
title: "Inference Etc."
format: html
embed-resources: true
---

```{r}
#| label: setup
#| message: false


library(tidyverse)
library(gt)
library(readxl)

Carseats <- ISLR2::Carseats
```

## Question 1

```{r}
mean(Carseats$Price,
     na.rm = TRUE)
```

The mean in this sample is 115.795 which is higher than what the data provider told us (110) we can investigate this with a visualization and t-test.

```{r}
ggplot(Carseats, aes(x = Price)) +
  geom_histogram() +
  geom_vline(xintercept = 110)
```
The distribution is more to the right of 110 than to the left, indicating 110 may not be the mean. 

We can figure this out using hypothesis testing:

Null Hypothesis: mu = 43 ($\mu = 43$)
Alt. Hypothesis: $H_a: \mu \ne 43$

```{r}
t.test(Carseats$Price,
       mu = 110)
```
The p-value<0.05 so we reject the null hypothesis.

In other words, the t-test tells us that there is strong evidence that the true mean is not equal to 110 as we were told. It suggests the sample mean is actually 115.795. 

## Question 2

```{r}
t.test(Carseats$Price, conf.level = 0.90)
```

The 90% confidence interval is [113.8432, 117.7468].
The sample mean is 115.795. 

```{r}
moe <- (117.7468 - 113.8432)/2
moe
```

The MOE is 1.9518

## Question 3

```{r}

t <- table(penguins$species, penguins$island)
addmargins(t)

```
## Question 4


